---
title: "Chapter 2 - Exercises"
description: |
  My attempt of the exericse questions from Chapter 2: Statistical Learning
author:
  - name: Kaustav Sen 
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Conceptual

## Question 7

Creating a function to calculate the "Euclidean distance" between two points.

```{r}
distance_euclidean <- function(x1, x2) {
  # assume that both x1 and x2 are vectors of the same length
  square_diff <- (x1 - x2) ^ 2
  sqrt(sum(square_diff))
}
```

Storing the data provided in the question as a `data.frame`

```{r}
data <- data.frame(x1 = c(0, 2, 0, 0, -1, 1),
                   x2 = c(3, 0, 1, 1, 0, 1),
                   x3 = c(0, 0, 3, 2, 1, 1),
                   y = c("Red", "Red", "Red", "Green", "Green", "Red"))
data
```

(a) Computing the Euclidean distance from the test point

```{r}
test_point <- c(0, 0, 0)
distances <- apply(data[, -4], 1, function(x) distance_euclidean(test_point, x))

distances
```

(b) When `K=1` the closest observation is observation 5. Thus, the prediction will be **Green**.

(c) When `K=3` the three closest observations will be 2, 5 and 6. Since, 2 and 6 are "Red", the prediction will be **Red**.

(d) If the Bayes decision boundary is highly non-linear, then we will expect the *best* value of K to be large. This is because we will need the model to be more flexible in order to capture the non-linear trend. This is only possible when K is high.

# Applied

## Question 8

(a) Loading in the `college.csv` data set and storing it in a variable called `college`.

```{r}
college <- read.csv("https://www.statlearning.com/s/College.csv")

head(college)
```

(b) Setting the row-names for the college data set and also using the `fix()` function

```{r}
rownames(college) <- college[,1]
# fix(college)
```

Eliminating the first column from the data.

```{r}
college <- college[, -1]
head(college)
```

(c) (i) Using the `summary()` function in order to get a numerical summary of the variables in the data.

```{r}
summary(college)
```

(ii) Using the `pairs()` function to plot a scatter plot for the first 10 columns in the data.

The first column is non-numeric, so ignored it from the subset.

```{r fig.width=8, fig.asp=1}
college_cols_10 <- college[, 2:10]

pairs(college_cols_10)
```

(iii) Creating a box-plot 

```{r fig.width=6, fig.asp=0.8}
plot(as.factor(college$Private), college$Outstate, 
     xlab = "Is Private?", 
     main = "How do Out-of-state tuition vary \nbetween Private and non-Private colleges?")
```

(iv) Creating a new qualitative variable, `Elite`

```{r}
Elite <- rep("No", nrow(college))
Elite[college$Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)
college <- data.frame(college, Elite)
```

```{r}
summary(college$Elite)
```

```{r fig.width=6, fig.asp=0.8}
plot(college$Elite, college$Outstate,
     xlab = "Is Elite?",
     ylab = "Out-of-state tuition",
     main = "Do Elite colleges charge more?")
```

(v) Creating some histograms

```{r fig.width=8, fig.asp=1}
par(mfrow = c(2, 2))
hist(college$Personal, breaks = 30, xlab = "Personal Spending", main = "")
hist(college$PhD, breaks = 30, xlab = "Percent of facultly with Ph.D.'s", main = "")
hist(college$S.F.Ratio, breaks = 30, xlab = "Student/faculty ratio", main = "")
hist(college$Grad.Rate, breaks = 30, xlab = "Graduation Rate", main = "")
```

(vi) Some further explorations of the data set...

```{r fig.width=6, fig.height=6}
# Using a scatter-plot to see if more expensive schools have a lower acceptance rate
acceptance_rate <- college$Accept / college$Apps
tuition_fees <- college$Outstate
par(mfrow = c(1,1))
plot(acceptance_rate, tuition_fees, main = "Do expensive schools have lower acceptance rates?")
```

```{r fig.width=6, fig.height=6}
plot(college$Grad.Rate, college$Expend, log = "y",
     xlab = "Graduation Rate", ylab = "Instructional Expenditure per student",
     main = "Does higher spending per student lead to higher\ngraduation rate?")
```

## Question 9

Reading in the `Auto` data and removing the `NA` values.

```{r}
Auto <- read.table("https://www.statlearning.com/s/Auto.data", 
                   header = TRUE, na.strings = "?")
Auto <- na.omit(Auto)
```

(a) Taking a look at the structure of the data set

```{r}
str(Auto)
```

The variables `name`, `cylinders` and `origin` are qualitative. Rest all of the other variables are quantitative in nature.

(b) Calculating the range of each quantitative variable.

```{r}
# Store the quantitative and qualitative variable names separately
qual_cols <- c("origin", "name", "cylinders")
quant_cols <- setdiff(names(Auto), qual_cols)

sapply(Auto[, quant_cols], range)
```

(c) Calculating the mean and standard deviation of each quantitative variable.

```{r}
output_summary_stats <- function(x) {
  list(mean = mean(x), standard_deviation = sd(x))
}
```

```{r}
sapply(Auto[, quant_cols], output_summary_stats)
```

(d) Sub-setting the data and then re-calculate the summary statistics.

```{r}
Auto_subset <- Auto[-c(10:85), quant_cols]
sapply(Auto_subset[, quant_cols], output_summary_stats)
```

(e) Exploring predictors using scatter-plots

```{r include=FALSE}
theme_custom <- function(base_family, base_size = 11) {
 theme_bw(base_family = base_family, base_size = base_size) %+replace%
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) 
}
```

```{r}
library(ggplot2)

plot_scatter_mpg <- function(var) {
 ggplot(Auto) +
  geom_point(aes({{ var }}, mpg, color = as.factor(cylinders)), 
             size = 2, alpha = 0.7) +
  guides(color = guide_legend("Number of cylinders", "top", title.hjust = 0.5)) +
  theme_custom(base_family = "Inter") 
}
```

```{r fig.width=10, fig.asp=0.8}
library(patchwork)

plot_scatter_mpg(displacement) + 
  plot_scatter_mpg(horsepower) + 
  plot_scatter_mpg(weight) +
  plot_scatter_mpg(acceleration) +
  plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "top")
```

(f) There seems to be a negative correlation between `mpg` and the following variables:

- `displacment`
- `horsepower` and
- `weight`

Plus, the number of cylinders also affects `mpg`. Further, there exists a slight positive correlation between `mpg` and `acceleration`. So, these varaibles can be used as a starting point for the modelling process.